Interactive Visualization of Hierarchically Structured Data

### 2017-03-15 treelapse.Rmd

```{r set_chunk_opts, echo = FALSE}
library("knitr")
opts_chunk$set(
  fig.align = "center",
  out.width = "250px",
  echo = FALSE,
  cache = TRUE
)
```

We introduce methods for visualization of data structured along trees,
especially hierarchically structured collections of time series. We
hope both to characterize generically useful techniques for
interactively visualizing hierarchical data and to offer practical
tools for implementing such displays. To this end, we identify
questions that often emerge when working with hierarchical data and
provide an R package to simplify their investigation [@ripley2001r].

@@

We have implemented our visualizations in D3, but encapsulated in an R package,
called treelapse, to facilitate rapid turnover from data preparation and
modeling to interactive exploration, and vice versa. Our code is open-source,
and available at \url{https://github.com/krisrs1128/treelapse}. We hope this
package encourages data analysts to work at the border between data modeling and
visualization, and more generally empowers a wider audience to apply less widely
known, but powerful, visualization ideas.

@@

### Hierarchically Clustering the Iris Data {#iris}

Each of the timebox tree and treebox examples presented so far have focused on
data with a clear time component. We note however that these techniques could
alternatively be applied to high-dimensional data, via the use of parallel
coordinates [@inselberg1991parallel]. The usual parallel coordinates challenges
remain, namely selecteing scales for and an ordering across the different
coordinates, but the linking and focus-plus-context ideas can still be employed
in this setting. Here we provide a simple implementation of this idea, using the
famous Iris data, inspired by the hierarchical clustering vignette in
[@galili2015dendextend].

The Iris data set is a collection of 150 measurements of four variables --
widths and lengths of Iris sepals and petals [@fisher1936use]. There are 50
samples for each of three species of Iris, \textit{setosa}, \textit{versicolor}
and \textit{virgnica}. As in [@galili2015dendextend], we hierarchical cluster
samples\footnote{We use complete linkage and the euclidean distance.}. Our
contribution is to display the resulting hierarchy together with a parallel
coordinates view of the raw data, see Figure \@ref(fig:iristimebox).

```{r iristimebox, fig.cap = "A toy application to the Iris data gives an example of how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views."}
include_graphics("figure/iris_timebox.png")
```

From Figure \@ref(fig:iristimebox), the fact that the highest split in the
hierarchical clustering distinguishes those samples with small petal lengths
from the rest. Upon inspecting the associated IDs, we find that this cluster
includes all the samples from setosa. The similarity between versicolor and
virginica is reflected in their proximity in the clustering. Further, applying a
treebox view to scan the subtrees associated with these two species reveals the
more subtle differences in sepal and petal lengths that can be used to
distinguish them.

>>>>

```{r set_chunk_opts, echo = FALSE}
library("knitr")
opts_chunk$set(
  fig.align = "center",
  out.width = "350px",
  echo = FALSE,
  cache = TRUE
)
```

We introduce methods for visualization of data structured along trees,
especially hierarchically structured collections of time series. We
hope both to characterize generically useful techniques for
interactively visualizing hierarchical data and to offer practical
tools for implementing such displays. To this end, we identify
questions that often emerge when working with hierarchical data and
provide an R package to simplify their investigation [@ihaka1996r].

@@

We have implemented our visualizations in D3, but encapsulated in an R package,
called treelapse, to facilitate rapid turnover from data preparation and
modeling to interactive exploration, and vice versa. Our code is open-source,
and available at \url{xxxxxxxxx}\footnote{Link removed in blinded manuscript,
but available to editor.}. We hope this package encourages data analysts to work
at the border between data modeling and visualization, and more generally
empowers a wider audience to apply less widely known, but powerful,
visualization ideas.

@@

### Hierarchically Clustering the Global Patterns Data {#global_patterns}

Each of the timebox tree and treebox examples presented so far have focused on
data with a clear time component. We note however that these techniques could
alternatively be applied to high-dimensional data, via the use of parallel
coordinates [@inselberg1991parallel]. The usual parallel coordinates challenges
remain, namely selecting scales for and an ordering across the different
coordinates, but the linking and focus-plus-context ideas can still be employed
in this setting. Here we provide an implementation of this idea on a dataset
comparing microbiomes across various ecological environments
[@caporaso2011global], which is publicly accessible through phyloseq
[@mcmurdie2013phyloseq].

The original Global Patterns data consists of 26 samples across 9 environments
(for example, freshwater and soil). In each site, there are counts across 19216
taxa -- to simplify visualization, we filter to the 500 most abundant taxa.

We hierarchically cluster these 26 samples based on the 500 most abundant taxa,
using complete linkage on the UniFrac distance. Figure \@ref(gptimebox) displays
the resulting hierarchy together with a parallel coordinates view of the
$\text{asinh}$ transformed taxa.

```{r gptimebox, fig.cap = "An application to the Global Patterns demonstrates how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views."}
include_graphics(c("figure/gp_cluster1.png", "figure/gp_cluster2.png"))
```

In Figure \@ref(fig:gptimebox), we compare two subclusters from the hierarchical
clustering tree, after zooming to a few of the microbes that distinguish between
the clusters. Upon revisiting the original data, it becomes clear that the
samples highlighted on the left come from freshwater samples, while those on the
right come from soil and skin, and looking up taxonomic groups associated with
the distinguishing microbes confirms this. For example, many of the microbes
with high abundances in the left figure come from order Oceanospirillales.
